---
title: "SARIMA_8regimes"
output: html_document
---
## Function for filtering based on regime

```{r}
plot_ccf_with_prewhitened_station <- function(df_he,
                                              df_x,
                                              regimes = c("R1", "R2", "R3", "R4", "R5", "R6", "R7", "R8"),
                                              time_col = "ts_utc",
                                              value_col = "Lufttemperatur",
                                              lag_max = 50) {
  
  library(dplyr)
  library(forecast)
  
  ## -----------------------------
  ## Prewhiten station (once)
  ## -----------------------------
  fit_x <- auto.arima(df_x[[value_col]])
  df_x$res <- residuals(fit_x)  # keep as ts
  
  ## -----------------------------
  ## Loop over regimes → separate plots
  ## -----------------------------
  for (reg in regimes) {
    
    df_he_R <- df_he %>% filter(regime == !!reg)
    t_R <- df_he_R[[time_col]]
    
    # Align station to regime times
    df_x_R <- df_x[df_x[[time_col]] %in% t_R, ]
    
    r <- ccf(df_x_R$res,
             df_he_R$res,
             lag.max = lag_max,
             na.action = na.pass,
             plot = FALSE)
    
    pos <- r$lag >= 0
    n <- sum(complete.cases(df_x_R$res, df_he_R$res))
    ci <- 1.96 / sqrt(n)
    
    plot(r$lag[pos], r$acf[pos], type = "h",
         xlab = "Lag",
         ylab = "CCF",
         main = paste("CCF –", reg, "(station prewhitened)"))
    
    abline(h = 0)
    abline(h = c(-ci, ci), lty = 2)
  }
}

```
## Loading data

```{r}
setwd("..")
source("Code/read_data.r")
```

```{r}
library(zoo)
library(forecast)
library(ggplot2)
library(dplyr)
```

```{r}
helsingborg_2024_2025 <- helsingborg %>%
  filter(
    ts_utc >= as.POSIXct("2024-01-01 00:00:00", tz = "UTC"),
    ts_utc <= as.POSIXct("2025-08-31 23:59:59", tz = "UTC")
  )
nidingen_2024_2025 <- nidingen %>%
  filter(
    ts_utc >= as.POSIXct("2024-01-01 00:00:00", tz = "UTC"),
    ts_utc <= as.POSIXct("2025-08-31 23:59:59", tz = "UTC")
  )
hörby_2024_2025 <- hörby %>%
  filter(
    ts_utc >= as.POSIXct("2024-01-01 00:00:00", tz = "UTC"),
    ts_utc <= as.POSIXct("2025-08-31 23:59:59", tz = "UTC")
  )
falsterbo_2024_2025 <- falsterbo %>%
  filter(
    ts_utc >= as.POSIXct("2024-01-01 00:00:00", tz = "UTC"),
    ts_utc <= as.POSIXct("2025-08-31 23:59:59", tz = "UTC")
  )
skillinge_2024_2025 <- skillinge %>%
  filter(
    ts_utc >= as.POSIXct("2024-01-01 00:00:00", tz = "UTC"),
    ts_utc <= as.POSIXct("2025-08-31 23:59:59", tz = "UTC")
  )
ullared_2024_2025 <- ullared %>%
  filter(
    ts_utc >= as.POSIXct("2024-01-01 00:00:00", tz = "UTC"),
    ts_utc <= as.POSIXct("2025-08-31 23:59:59", tz = "UTC")
  )
ljungby_2024_2025 <- ljungby %>%
  filter(
    ts_utc >= as.POSIXct("2024-01-01 00:00:00", tz = "UTC"),
    ts_utc <= as.POSIXct("2025-08-31 23:59:59", tz = "UTC")
  )
hästveda_2024_2025 <- hästveda %>%
  filter(
    ts_utc >= as.POSIXct("2024-01-01 00:00:00", tz = "UTC"),
    ts_utc <= as.POSIXct("2025-08-31 23:59:59", tz = "UTC")
  )
```

## "Stationary" period

```{r}
start <- 3200
end <- 5900
df_he <- helsingborg_2024_2025[start:end, ] # center
df_ni <- nidingen_2024_2025[start:end, ] # north west
df_fa <- falsterbo_2024_2025[start:end, ] # south
df_hä <- hästveda_2024_2025[start:end, ] # east north east
df_lj <- ljungby_2024_2025[start:end, ] # north east
df_ul <- ullared_2024_2025[start:end, ] # north
df_sk <- skillinge_2024_2025[start:end, ] # south east
df_hö <- hörby_2024_2025[start:end, ] # east
```

## Training and testing split

```{r}
h <- 500
freq <- 24
# full series from your dataframe
y_all <- ts(helsingborg_2024_2025$Lufttemperatur, frequency = 24)

# convert row indices to ts "time" values
t_start <- 1 + (start - 1) / freq
t_end   <- 1 + (end   - 1) / freq
t_test_start <- 1 + (end) / freq          # row end + 1
t_test_end   <- 1 + (end + h - 1) / freq  # row end + h

# train/test as ts
y_train <- window(y_all, start = t_start, end = t_end)
y_test  <- window(y_all, start = t_test_start, end = t_test_end)

# additional test set roughly 6 months forward in time

gap <- 4000  

t_test1_start <- t_test_start + gap / freq
t_test1_end   <- t_test1_start + (h - 1) / freq

y_test1 <- window(y_all, start = t_test1_start, end = t_test1_end)

```

## Fixed and non-fixed parameters in p/q polynomials

```{r}
p <- rep(0, 18)
p[1] <- NA
p[2] <- NA
p[11] <- 0
p[18] <- NA

q <- rep(0, 23)
q[2] <- NA
q[20] <- NA
q[23] <- NA
```

## Seasonal ARIMA model:

```{r}
fit <- Arima(y_train,
                 order = c(18,0,23),
                fixed = c(p,q,NA),
                 seasonal = c(0,1,1))
```

## Adding residuals to df

```{r}
df_he$res <- fit$residuals
```

## 8 wind regimes

```{r}
wind_regime <- function(wd, ws, rot = 45) {
  wd <- (wd - rot) %% 360
  out <- rep(NA_character_, length(wd))

  # Only assign R1–R4 when wind speed is not zero, when zero, set regime to R5
  idx <- !is.na(ws) & ws != 0 & !is.na(wd)
 
  out[!is.na(ws) & ws == 0] <- "R0"
  out[idx & wd >   0 & wd <=  45] <- "R1"
  out[idx & wd >  45 & wd <= 90] <- "R2"
  out[idx & wd >  90 & wd <= 135] <- "R3"
  out[idx & wd >  135 & wd <= 180] <- "R4"
  out[idx & wd > 180 & wd <= 225] <- "R5"
  out[idx & wd > 225 & wd <  270] <- "R6"
  out[idx & wd > 270 & wd <  315] <- "R7"
  out[idx & wd > 315 & wd <  360] <- "R8"
  out[idx & wd == 0]              <- "R8"

  factor(out, levels = c("R0","R1","R2", "R3","R4","R5", "R6", "R7", "R8"))

}

df_he$regime <- wind_regime(df_he$Vindriktning, df_he$Vindhastighet, rot = 45)
```

## Add danish weather station

```{r}
setwd("..")
source("Code/DenmarkFunctions.r")
```

```{r}
# Get available parameters for the station
params <- c("temp_dry","wind_dir","wind_speed")

start_time <- format(as.POSIXct(df_he$ts_utc[1], tz = "UTC"),
       "%Y-%m-%dT%H:%M:%SZ")
end_time <- format(as.POSIXct(df_he$ts_utc[nrow(df_he)], tz = "UTC"),
       "%Y-%m-%dT%H:%M:%SZ")
```

## Loading wanted stations
```{r}
stations <- c(
  kb = "06180", # Kobenhavn lufthavn
  ro = "06170", # Roskilde Lufthavn
  gn = "06049", # Gniben
  sl = "06073", # Sletterhage Fyr
  an = "06079"  # Anholt
)

df_list <- vector("list", length(stations))
names(df_list) <- names(stations)

missing_list <- vector("list", length(stations))
names(missing_list) <- names(stations)

for (nm in names(stations)) {
  station_id <- stations[[nm]]

  observations <- get_observations_for_station(
    station_id = station_id,
    parameters = params,
    start_dt   = start_time,
    end_dt     = end_time,
    limit      = 20000
  )

  full_seq <- seq(
    from = min(observations$observed),
    to   = max(observations$observed),
    by   = "10 min"
  )

  missing_times <- full_seq[!full_seq %in% observations$observed]
  missing_list[[nm]] <- missing_times

  df <- process_hourly_data(observations) %>%
    rename(
      Lufttemperatur = temp_dry,
      Vindriktning   = wind_dir,
      Vindhastighet  = wind_speed,
      ts_utc         = observed
    )

  fit_x <- forecast::auto.arima(df$Lufttemperatur)
  df$res <- residuals(fit_x)

  df_list[[nm]] <- df
}

# If you want:
df_kb <- df_list$kb
df_ro <- df_list$ro
df_gn <- df_list$gn
df_sl <- df_list$sl
df_an <- df_list$an
```

```{r}
table(df_he$regime, useNA = "ifany")
```


### R1
## Hästveda
```{r}
plot_ccf_with_prewhitened_station(
  df_he = df_he,
  df_x  = df_hä,
  regimes = "R1"
)
```

### R2
## Hörby
```{r}
plot_ccf_with_prewhitened_station(
  df_he = df_he,
  df_x  = df_hö,
  regimes = "R2"
)
```

### R3
## Skillinge
```{r}
plot_ccf_with_prewhitened_station(
  df_he = df_he,
  df_x  = df_sk,
  regimes = "R3"
)
```

### R4
## Falsterbo
```{r}
plot_ccf_with_prewhitened_station(
  df_he = df_he,
  df_x  = df_fa,
  regimes = "R4"
)
```
## Köbenhavn lufthavn
```{r}
plot_ccf_with_prewhitened_station(
  df_he = df_he,
  df_x  = df_kb,
  regimes = "R4"
)
```
## Roskilde Lufthavn
```{r}
plot_ccf_with_prewhitened_station(
  df_he = df_he,
  df_x  = df_ro,
  regimes = "R4"
)
```

### R5
## Roskilde lufthavn
```{r}
plot_ccf_with_prewhitened_station(
  df_he = df_he,
  df_x  = df_ro,
  regimes = "R5"
)
```
## Gniben
```{r}
plot_ccf_with_prewhitened_station(
  df_he = df_he,
  df_x  = df_gn,
  regimes = "R5"
)
```
## Sletterhage fyr
```{r}
plot_ccf_with_prewhitened_station(
  df_he = df_he,
  df_x  = df_sl,
  regimes = "R5"
)
```

### R6
## Anholt
```{r}
plot_ccf_with_prewhitened_station(
  df_he = df_he,
  df_x  = df_an,
  regimes = "R6"
)
```

### R7
## Nidingen
```{r}
plot_ccf_with_prewhitened_station(
  df_he = df_he,
  df_x  = df_ni,
  regimes = "R7"
)
```
## Ullared
```{r}
plot_ccf_with_prewhitened_station(
  df_he = df_he,
  df_x  = df_ul,
  regimes = "R7"
)
```

### R8
##Ljungby
```{r}
plot_ccf_with_prewhitened_station(
  df_he = df_he,
  df_x  = df_lj,
  regimes = "R8"
)
```
